A Configurable Library for Generating and Manipulating Maze Datasets

### 2025-03-25 docs/paper/paper.tex
modify paper, recompile

\documentclass[10pt,a4paper,onecolumn]{article}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage[rgb,svgnames]{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{tikz}
\usepackage[pdfa]{hyperref}
\usepackage{hyperxmp}
\hypersetup{%
    unicode=true,
    pdfapart=3,
    pdfaconformance=B,
    pdftitle={A Configurable Library for Generating and Manipulating
Maze Datasets},
    pdfauthor={Michael Igorevich Ivanitskiy, Aaron Sandoval, Alex F.
Spies, Rusheb Shah, Brandon Knutson, Cecilia Diniz Behn, Samy Wu Fung},
    pdfpublication={Journal of Open Source Software},
    pdfpublisher={Open Journals},
    pdfissn={2475-9066},
    pdfpubtype={journal},
    pdfvolumenum={},
    pdfissuenum={},
    pdfdoi={N/A},
    pdfcopyright={Copyright (c) 1970, Michael Igorevich Ivanitskiy,
Aaron Sandoval, Alex F. Spies, Rusheb Shah, Brandon Knutson, Cecilia
Diniz Behn, Samy Wu Fung},
    pdflicenseurl={http://creativecommons.org/licenses/by/4.0/},
    colorlinks=true,
    linkcolor=[rgb]{0.0, 0.5, 1.0},
    citecolor=Blue,
    urlcolor=[rgb]{0.0, 0.5, 1.0},
    breaklinks=true
}
% https://tex.stackexchange.com/a/535849
% Create an OutputIntent in order to correctly specify colours
\immediate\pdfobj stream attr{/N 3} file{sRGB.icc}
\pdfcatalog{%
  /OutputIntents [
    <<
      /Type /OutputIntent
      /S /GTS_PDFA1
      /DestOutputProfile \the\pdflastobj\space 0 R
      /OutputConditionIdentifier (sRGB)
      /Info (sRGB)
    >>
  ]
}
\pdfvariable omitcidset=1
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{tcolorbox}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{seqsplit}
\usepackage{xstring}

@@

\author[1%
%
\ensuremath\mathparagraph]{Michael Igorevich Ivanitskiy%
  \,\orcidlink{0000-0002-4213-4993}\,%
}
\author[%
%
]{Aaron Sandoval%
}
\author[2%
%
]{Alex F. Spies%
  \,\orcidlink{0000-0002-8708-1530}\,%
}
\author[%
%
]{Rusheb Shah%
  \,\orcidlink{0009-0008-3058-0217}\,%
}
\author[%
%
]{Brandon Knutson%
}
\author[1%
%
]{Cecilia Diniz Behn%
}
\author[1%
%
]{Samy Wu Fung%
}

@@

Understanding how machine learning models respond to distributional
shifts and different input representations is a key research challenge.
Mazes serve as an excellent testbed for this problem due to the degree
of control we have over quantifiable distributional shifts. To enable
systematic investigations of model behavior on maze tasks, we present
\texttt{maze-dataset}, a comprehensive library for generating,
processing, visualizing, and tokenizing datasets consisting of
maze-solving tasks. With this library, researchers can easily create
datasets and have extensive control over the generation algorithm used,
the parameters fed to the algorithm of choice, and the filters that
generated mazes must satisfy. Furthermore, this library supports
multiple output formats, including rasterized and a large number of
text-based formats, catering to convolutional neural networks and
autoregressive transformer models. These formats, along with tools for
visualizing and converting between them, ensure versatility and
adaptability in research applications.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Out-of-distribution generalization is a critical challenge in modern
machine learning (ML) research. For interpretability and behavioral
research in this area, training on algorithmic tasks offers benefits by
allowing systematic data generation and task decomposition, as well as
simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we have found
that existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) do not support flexible maze generation
algorithms that provide fine-grained control of generation parameters
and the ability to easily transform between multiple representations of
the mazes (Images, Textual, Tokenized) for training and testing models.

This work aims to facilitate deeper research into generalization and
interpretability by addressing these limitations. We introduce
\href{https://github.com/understanding-search/maze-dataset}{\texttt{maze-dataset}},
an accessible Python package
(\protect\hyperlink{ref-maze-dataset-github}{M. I. Ivanitskiy et al.,
2023a}). This package offers flexible configuration options for maze
dataset generation, allowing users to select from a range of algorithms
and adjust corresponding parameters (Section
\protect\hyperlink{generation}{Generation}). Furthermore, this package
supports various output formats tailored to different ML architectures
(Section \protect\hyperlink{visual-output-formats}{Visual Output
Formats}, Section \protect\hyperlink{tokenized-output-formats}{Tokenized
Output Formats}).

TODO: links to docs, not just github

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth,height=\textheight]{figures/demo-dataset.pdf}
\caption{Example mazes from various algorithms. Left to right:
randomized depth-first search (RDFS), RDFS without forks, constrained
RDFS, Wilson's (\protect\hyperlink{ref-wilson}{Wilson, 1996}), RDFS with
percolation (p=0.1), RDFS with percolation (p=0.4), random stack RDFS.}
\end{figure}

\hypertarget{generation}{%
\section{Generation and Usage}\label{generation}}

Our package can be installed from
\href{https://pypi.org/project/maze-dataset/}{PyPi} via
\texttt{pip\ install\ maze-dataset}, or directly from the
\href{https://github.com/understanding-search/maze-dataset}{git
repository} (\protect\hyperlink{ref-maze-dataset-github}{M. I.
Ivanitskiy et al., 2023a}).

@@

When initializing mazes, further configuration options can be specified
through the \texttt{from\_config()} factory method as necessary. Options
include 1) whether to generate the dataset during runtime or load an
existing dataset, 2) if and how to parallelize generation, and 3) where
to store the generated dataset. Full documentation of configuration
options is available in our repository
(\protect\hyperlink{ref-maze-dataset-github}{M. I. Ivanitskiy et al.,
2023a}). Available maze generation algorithms are static methods of the
\texttt{LatticeMazeGenerators} class and include the following:

@@

Examples of usage of this dataset to train autoregressive transformers
can be found in our \texttt{maze-transformer} library
(\protect\hyperlink{ref-maze-transformer-github}{M. I. Ivanitskiy et
al., 2023b}). Other tokenization and vocabulary schemes are also
included, such as representing each coordinate as a pair of \(i,j\)
index tokens.

@@

We refer to our GitHub repository
(\protect\hyperlink{ref-maze-dataset-github}{M. I. Ivanitskiy et al.,
2023a}) for documentation and up-to-date implementation details.

@@

\hypertarget{relation-work}{%
\subsection{Relation to Existing Works}\label{relation-work}}

As mentioned in the introduction, a multitude of public and open-source
software packages exist for generating mazes
(\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). However, our package provides more flexibility
and efficiency in the following ways:

\begin{itemize}
\tightlist
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). This metadata is loaded when the dataset is
  retrieved from disk and reduces the complexity of discerning the
  parameters under which a dataset was created.
\item
  Prior works provide maze datasets in only a rasterized format, which
  is not suitable for training autoregressive text-based transformer
  models. As discussed in Section
  \protect\hyperlink{visual-output-formats}{Visual Output Formats} and
  Section \protect\hyperlink{tokenized-output-formats}{Tokenized Output
  Formats}, our package provides these different formats natively.
\item
  Our package provides a selection of maze generation algorithms, which
  all write to a single unified format. All output formats are
  reversible, and operate to and from this unified format.
\end{itemize}

As mentioned in Section \protect\hyperlink{training}{Training}, we also
include the \texttt{RasterizedMazeDataset} class in our codebase, which
can exactly mimic the outputs provided in
\texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}). Our \texttt{as\_ascii()} method
provides a format similar to that used in
(\protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023}). The text format
provided by \texttt{as\_tokens()} is similar to that of
(\protect\hyperlink{ref-eval-LLM-graphs}{Liu \& Wu, 2023}), but provides
a custom tokenization scheme.

@@

@@

The
\href{https://github.com/understanding-search/maze-dataset}{\texttt{maze-dataset}}
library (\protect\hyperlink{ref-maze-dataset-github}{M. I. Ivanitskiy et
al., 2023a}) introduced in this paper provides a flexible and extensible
toolkit for generating, processing, and analyzing maze datasets. By
supporting various procedural generation algorithms and conversion
utilities, it enables the creation of mazes with customizable properties
to suit diverse research needs. Planned improvements to the
\texttt{maze-dataset} include adding more generation algorithms (such as
Prim's algorithm (\protect\hyperlink{ref-dijkstra-prim}{Dijkstra, 1959};
\protect\hyperlink{ref-jarnik-prim}{Jarnık, 1930};
\protect\hyperlink{ref-prim}{Prim, 1957}) and Kruskal's algorithm
(\protect\hyperlink{ref-kruskal}{Kruskal, 1956}), among others
(\protect\hyperlink{ref-mazegen_analysis}{Gabrovšek, 2019})), adding the
ability to augment a maze with an adjacency list to add ``shortcuts'' to
the maze, and resolving certain limitations detailed in Section
\protect\hyperlink{limitations}{Limitations}. Future work will make
extensive use of this library to study interpretability and
out-of-distribution generalization in autoregressive transformers
(\protect\hyperlink{ref-maze-transformer-github}{M. I. Ivanitskiy et
al., 2023b}), recurrent convolutional neural networks
(\protect\hyperlink{ref-deepthinking}{Schwarzschild, Borgnia, Gupta,
Huang, et al., 2021}), and implicit networks
(\protect\hyperlink{ref-fung2022jfb}{Fung et al., 2022};
\protect\hyperlink{ref-mckenzie2023faster}{McKenzie et al., 2023}).

@@

This work was partially supported by and many of the authors were
brought together by AI Safety Camp and AI Safety Support. This work was
partially funded by National Science Foundation awards DMS-2110745 and
DMS-2309810. We thank the Mines Optimization and Deep Learning group
(MODL) for fruitful discussions. We also thank Michael Rosenberg for
advice on Finite State Transducers.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0.5}
\leavevmode\vadjust pre{\hypertarget{ref-cobbe2019procgen}{}}%
Cobbe, K., Hesse, C., Hilton, J., \& Schulman, J. (2019). Leveraging
procedural generation to benchmark reinforcement learning. \emph{arXiv
Preprint arXiv:1912.01588}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-maze-dataset-github}{}}%
Ivanitskiy, M. I., Shah, R., Spies, A. F., Räuker, T., Valentine, D.,
Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023a). \emph{Maze
dataset}. \url{https://github.com/understanding-search/maze-dataset}

\leavevmode\vadjust pre{\hypertarget{ref-maze-transformer-github}{}}%
Ivanitskiy, M. I., Shah, R., Spies, A. F., Räuker, T., Valentine, D.,
Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023b). \emph{Maze
transformer interpretability}.
\url{https://github.com/understanding-search/maze-transformer}

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-mckenzie2023faster}{}}%
McKenzie, D., Fung, S. W., \& Heaton, H. (2023). Faster
predict-and-optimize with three-operator splitting. \emph{arXiv Preprint
arXiv:2301.13395}.

@@

@@

>>>>

\documentclass[10pt,a4paper,onecolumn]{article}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage[rgb,svgnames]{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{tikz}
\usepackage[pdfa]{hyperref}
\usepackage{hyperxmp}
\hypersetup{%
    unicode=true,
    pdfapart=3,
    pdfaconformance=B,
    pdftitle={A Configurable Library for Generating and Manipulating
Maze Datasets},
    pdfauthor={Michael Igorevich Ivanitskiy, Aaron Sandoval, Alex F.
Spies, Rusheb Shah, Brandon Knutson, , Cecilia Diniz Behn, Samy Wu
Fung},
    pdfpublication={Journal of Open Source Software},
    pdfpublisher={Open Journals},
    pdfissn={2475-9066},
    pdfpubtype={journal},
    pdfvolumenum={},
    pdfissuenum={},
    pdfdoi={N/A},
    pdfcopyright={Copyright (c) 1970, Michael Igorevich Ivanitskiy,
Aaron Sandoval, Alex F. Spies, Rusheb Shah, Brandon Knutson, , Cecilia
Diniz Behn, Samy Wu Fung},
    pdflicenseurl={http://creativecommons.org/licenses/by/4.0/},
    colorlinks=true,
    linkcolor=[rgb]{0.0, 0.5, 1.0},
    citecolor=Blue,
    urlcolor=[rgb]{0.0, 0.5, 1.0},
    breaklinks=true
}
% https://tex.stackexchange.com/a/535849
% Create an OutputIntent in order to correctly specify colours
\immediate\pdfobj stream attr{/N 3} file{sRGB.icc}
\pdfcatalog{%
  /OutputIntents [
    <<
      /Type /OutputIntent
      /S /GTS_PDFA1
      /DestOutputProfile \the\pdflastobj\space 0 R
      /OutputConditionIdentifier (sRGB)
      /Info (sRGB)
    >>
  ]
}
\pdfvariable omitcidset=1
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{tcolorbox}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{seqsplit}
\usepackage{xstring}

@@

\author[1%
%
\ensuremath\mathparagraph]{Michael Igorevich Ivanitskiy%
  \,\orcidlink{0000-0002-4213-4993}\,%
}
\author[%
%
]{Aaron Sandoval%
}
\author[2%
%
]{Alex F. Spies%
  \,\orcidlink{0000-0002-8708-1530}\,%
}
\author[%
%
]{Rusheb Shah%
  \,\orcidlink{0009-0008-3058-0217}\,%
}
\author[%
%
]{Brandon Knutson%
}
\author[1%
%
]{%
  \,\orcidlink{0009-0004-8413-0239}\,%
}
\author[1%
%
]{Cecilia Diniz Behn%
}
\author[1%
%
]{Samy Wu Fung%
}

@@

Solving mazes is a classic problem in computer science and artificial
intelligence, and humans have been constructing mazes for thousands of
years. Although finding the shortest path through a maze is a solved
problem, this makes it an excellent testbed for studying how machine
learning algorithms solve problems and represent spatial information. In
this paper, we introduce \texttt{maze-dataset}, a Python library for
generating, processing, and visualizing datasets of mazes. This library
supports a variety of maze generation algorithms providing both mazes
with loops and ``perfect'' mazes without them. These generation
algorithms can be configured with various parameters, and the resulting
mazes can be filtered to satisfy a desired property. Also provided are
tools for converting mazes to and from various formats, such as
rasterized images and tokenized text sequences, making it suitable for
training or evaluating convolutional neural networks and autoregressive
transformer models, as well as various visualization tools. As well as
providing a simple interface for generating, storing, and loading these
datasets, \texttt{maze-dataset} is extensively tested, type hinted,
benchmarked, and documented.

\textbf{TODO: mention the examples, put in a figure
https://understanding-search.github.io/maze-dataset/examples/maze\_examples.html}

\hypertarget{statement-of-need}{%
\section{Statement of Need}\label{statement-of-need}}

The generation of mazes with a given algorithm is not inherently a
complex task, but the ability to seamlessly switch out algorithms,
modify algorithm parameters, or filter by desired properties all while
preserving the ability to convert between different representations of
the maze is not trivial. This library aims to greatly streamline the
process of generating and working with datasets of mazes that can be
described as subgraphs of an \(n \times n\) lattice with boolean
connections and, optionally, start and end points that are nodes in the
graph. Furthermore, we place emphasis on a wide variety of possible text
output formats aimed at evaluating the spatial reasoning capabilities of
Large Language Models and other text-based transformer models.

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we have found
that existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) do not support flexible maze generation
algorithms that provide fine-grained control of generation parameters
and the ability to easily transform between multiple representations of
the mazes (Images, Textual, Tokenized) for training and testing models.

\hypertarget{related-works}{%
\subsection{Related Works}\label{related-works}}

A multitude of public and open-source software packages exist for
generating mazes (\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). However, nearly all of these packages generate
and store mazes in a form that is not optimized for storage space or,
more importantly, computer readability. The mazes produces by other
packages are usually rasterized or in some form of image, rather than
the underlying graph structure, and this makes it difficult to work with
these datasets.

\begin{itemize}
\item
  Most prior works provide mazes in some kind of image or raster format,
  which is not suitable for training autoregressive text-based
  transformer models -- a key usage case this work seeks to enable.
  However, we still provide a variety of similar output formats:

  \begin{itemize}
  \tightlist
  \item
    we also include the \texttt{RasterizedMazeDataset} class, utilizing
    \texttt{as\_pixels()}, in our codebase, which can exactly mimic the
    outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of
    (\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019}).
  \item
    Our \texttt{as\_ascii()} method provides a format similar to that
    used in Oppenheim
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{2018}).
  \item
    Our \texttt{MazePlot} class provides a feature-rich plotting utility
    with support for multiple paths, heatmaps over positions, and more.
    This is similar to the outputs of Ehsan
    (\protect\hyperlink{ref-gh_Ehsan_2022}{2022})
  \end{itemize}
\item
  The text format provided by \texttt{SolvedMaze(...).as\_tokens()} is
  similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}), but provides over 5.8 million unique formats for converting
  mazes to a text stream. We maintain a single underlying format,
  meaning that the same maze can be turned into a variety of text
  streams to assess how the precise format of the text stream affects
  the model.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). This metadata is loaded when the dataset is
  retrieved from disk and makes it simple to understand how exactly each
  maze was generated. As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. Directly storing adjacency matrices is also inefficient
  as subgraphs of the lattice are sparse. Storing adjacency lists can be
  efficient, but comes with a higher lookup cost and possible high
  comparison cost. We use a simple, efficient representation of mazes
  that is optimized for subgraphs of a \(d\)-dimensional finite lattice
  that we do not believe is used in any existing maze generation
  package.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

\hypertarget{features-and-implementation}{%
\section{Features and
Implementation}\label{features-and-implementation}}

\hypertarget{generation}{%
\subsection{Generation and Usage}\label{generation}}

\textbf{TODO: JOSS says don't document the API itself. we should talk
about the capabilities, but just rewrite this to link to the
documentation.}

Our package can be installed from
\href{https://pypi.org/project/maze-dataset/}{PyPi} via
\texttt{pip\ install\ maze-dataset}, or directly from the
\href{https://github.com/understanding-search/maze-dataset}{git
repository} (\protect\hyperlink{ref-maze-dataset-github}{Michael I.
Ivanitskiy et al., 2023a}).

@@

When initializing mazes, further configuration options can be specified
through the \texttt{from\_config()} factory method as necessary. Options
include 1) whether to generate the dataset during runtime or load an
existing dataset, 2) if and how to parallelize generation, and 3) where
to store the generated dataset. Full documentation of configuration
options is available in our repository
(\protect\hyperlink{ref-maze-dataset-github}{Michael I. Ivanitskiy et
al., 2023a}). Available maze generation algorithms are static methods of
the \texttt{LatticeMazeGenerators} class and include the following:

@@

Examples of usage of this dataset to train autoregressive transformers
can be found in our \texttt{maze-transformer} library
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}). Other tokenization and vocabulary schemes are also
included, such as representing each coordinate as a pair of \(i,j\)
index tokens.

@@

We refer to our GitHub repository
(\protect\hyperlink{ref-maze-dataset-github}{Michael I. Ivanitskiy et
al., 2023a}) for documentation and up-to-date implementation details.

@@

@@

\hypertarget{usage-in-research}{%
\section{Usage in Research}\label{usage-in-research}}

This package was originally built for the needs of the
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}) project, which aims to investigate spatial planning and
world models in autoregressive transformer models trained on mazes Spies
et al. (\protect\hyperlink{ref-spies2024causalworldmodels}{2024}). This
project has also adapted itself to be useful for work on understanding
the mechanisms by which recurrent convolutional and implicit networks
(\protect\hyperlink{ref-fung2022jfb}{Fung et al., 2022}) solve mazes
given a rasterized view
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), and for this we match the output format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}).

This package has also been utilized in work by other groups:

\begin{itemize}
\item
  (\protect\hyperlink{ref-nolte2024multistep}{Nolte et al., 2024}) use
  \texttt{maze-dataset} to compare the effectiveness of transformers
  trained with the MLM-\(\mathcal{U}\)
  (\protect\hyperlink{ref-MLMU-kitouni2024factorization}{Kitouni et al.,
  2024}) multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  (\protect\hyperlink{ref-wang2024imperative}{Wang et al., 2024}) and
  (\protect\hyperlink{ref-chen2024iaimperative}{Chen et al., 2024}) use
  \texttt{maze-dataset} to study the effectiveness of imperative
  learning
\end{itemize}

@@

The
\href{https://github.com/understanding-search/maze-dataset}{\texttt{maze-dataset}}
library (\protect\hyperlink{ref-maze-dataset-github}{Michael I.
Ivanitskiy et al., 2023a}) introduced in this paper provides a flexible
and extensible toolkit for generating, processing, and analyzing maze
datasets. By supporting various procedural generation algorithms and
conversion utilities, it enables the creation of mazes with customizable
properties to suit diverse research needs. Planned improvements to the
\texttt{maze-dataset} include adding more generation algorithms (such as
Prim's algorithm (\protect\hyperlink{ref-dijkstra-prim}{Dijkstra, 1959};
\protect\hyperlink{ref-jarnik-prim}{Jarnık, 1930};
\protect\hyperlink{ref-prim}{Prim, 1957}) and Kruskal's algorithm
(\protect\hyperlink{ref-kruskal}{Kruskal, 1956}), among others
(\protect\hyperlink{ref-mazegen_analysis}{Gabrovšek, 2019})), adding the
ability to augment a maze with an adjacency list to add ``shortcuts'' to
the maze, and resolving certain limitations detailed in Section
\protect\hyperlink{limitations}{Limitations}.

@@

This work was partially supported by and many of the authors were
brought together by AI Safety Camp and AI Safety Support. This work was
partially funded by National Science Foundation awards DMS-2110745 and
DMS-2309810. We are also grateful to LTFF and FAR Labs for hosting three
of the authors for a Residency Visit, and to various members of FAR's
technical staff for their advice. We thank the Mines Optimization and
Deep Learning group (MODL) for fruitful discussions. We also thank
Michael Rosenberg for recommending the usage of Finite State Transducers
for storing tokenizer validation information.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0.5}
\leavevmode\vadjust pre{\hypertarget{ref-mazegenerator-net}{}}%
Alance AB. (2019). \emph{Maze generator}.
\url{http://www.mazegenerator.net}.

\leavevmode\vadjust pre{\hypertarget{ref-ayaz2008maze}{}}%
Ayaz, H., Allen, S. L., Platek, S. M., \& Onaral, B. (2008). Maze suite
1.0: A complete set of tools to prepare, present, and analyze
navigational and spatial cognitive neuroscience experiments.
\emph{Behavior Research Methods}, \emph{40}, 353--359.

\leavevmode\vadjust pre{\hypertarget{ref-chen2024iaimperative}{}}%
Chen, X., Yang, F., \& Wang, C. (2024). iA \(\^{}*\): Imperative
learning-based a \(\^{}*\) search for pathfinding. \emph{arXiv Preprint
arXiv:2403.15870}.

\leavevmode\vadjust pre{\hypertarget{ref-cobbe2019procgen}{}}%
Cobbe, K., Hesse, C., Hilton, J., \& Schulman, J. (2019). Leveraging
procedural generation to benchmark reinforcement learning. \emph{arXiv
Preprint arXiv:1912.01588}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mathematica-maze}{}}%
Guo, C., Barthelet, L., \& Morris, R. (2011). \emph{Maze generator and
solver}. Wolfram Demonstrations Project,
\url{https://demonstrations.wolfram.com/MazeGeneratorAndSolver/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-maze-dataset-github}{}}%
Ivanitskiy, Michael I., Shah, R., Spies, A. F., Räuker, T., Valentine,
D., Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023a).
\emph{Maze dataset}.
\url{https://github.com/understanding-search/maze-dataset}

\leavevmode\vadjust pre{\hypertarget{ref-maze-transformer-github}{}}%
Ivanitskiy, Michael I., Shah, R., Spies, A. F., Räuker, T., Valentine,
D., Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023b).
\emph{Maze transformer interpretability}.
\url{https://github.com/understanding-search/maze-transformer}

\leavevmode\vadjust pre{\hypertarget{ref-ivanitskiy2023structuredworldreps}{}}%
Ivanitskiy, Michael Igorevich, Spies, A. F., Räuker, T., Corlouer, G.,
Mathwin, C., Quirke, L., Rager, C., Shah, R., Valentine, D., Behn, C.
D., \& others. (2023). Structured world representations in maze-solving
transformers. \emph{arXiv Preprint arXiv:2312.02566}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-MLMU-kitouni2024factorization}{}}%
Kitouni, O., Nolte, N. S., Williams, A., Rabbat, M., Bouchacourt, D., \&
Ibrahim, M. (2024). The factorization curse: Which tokens you predict
underlie the reversal curse and more. \emph{Advances in Neural
Information Processing Systems}, \emph{37}, 112329--112355.

\leavevmode\vadjust pre{\hypertarget{ref-knutson2024logicalextrapolation}{}}%
Knutson, B., Rabeendran, A. C., Ivanitskiy, M., Pettyjohn, J.,
Diniz-Behn, C., Fung, S. W., \& McKenzie, D. (2024). On logical
extrapolation for mazes with recurrent and implicit networks.
\emph{arXiv Preprint arXiv:2410.03020}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mdl-suite}{}}%
Nag, A. (2020). MDL suite: A language, generator and compiler for
describing mazes. \emph{Journal of Open Source Software}, \emph{5}(46),
1815.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nolte2024multistep}{}}%
Nolte, N., Kitouni, O., Williams, A., Rabbat, M., \& Ibrahim, M. (2024).
Transformers can navigate mazes with multi-step prediction. \emph{arXiv
Preprint arXiv:2412.05117}.

\leavevmode\vadjust pre{\hypertarget{ref-gh-oppenheimj2018maze}{}}%
Oppenheim, J. (2018). \emph{Maze-generator: Generate a random maze
represented as a 2D array using depth-first search}.
\url{https://github.com/oppenheimj/maze-generator/}; GitHub.

@@

\leavevmode\vadjust pre{\hypertarget{ref-spies2024causalworldmodels}{}}%
Spies, A. F., Edwards, W., Ivanitskiy, M. I., Skapars, A., Räuker, T.,
Inoue, K., Russo, A., \& Shanahan, M. (2024). Transformers use causal
world models in maze-solving tasks. \emph{arXiv Preprint
arXiv:2412.11867}.

\leavevmode\vadjust pre{\hypertarget{ref-wang2024imperative}{}}%
Wang, C., Ji, K., Geng, J., Ren, Z., Fu, T., Yang, F., Guo, Y., He, H.,
Chen, X., Zhan, Z., \& others. (2024). Imperative learning: A
self-supervised neural-symbolic learning framework for robot autonomy.
\emph{arXiv Preprint arXiv:2406.16087}.
